---
title: DuckDB SQL
subtitle: Use the same queries from R, Python, Julia...
execute:
  freeze: auto
  cache: true
---

## What is a Database?

You can think of a database as a collection of tables (or dataframes). In general, the benefit of databases is that they can handle very large amounts of data - data which not fit in your computer's RAM. As such, most databases are stored on hard disk, from where data is read and written to. 

There are many different forms of databases. We have several so-called database management systems (DBMS), which allow us to interact with each of them. Unhelpfully, each different flavor of database has a slightly different dialect of a what is called the *structured query language*, **SQL** for short. Helpfully, on the other hand, there are several packages out there which help us keeping the effort manageable by translating our queries. We will see a few of those here, and in different driving languages (i.e. `R`, `python` and `julia`).

One final important feature of different databases is how (and where) they actually run. The more traditional solutions exist as server-client models, where you would start on a machine (could be your own) a database server, which interacts with the persistent storage (e.g. with your hard disk) to read and write data, and a client process, where you formulate so-called _queries_, and get back the results from the server. In some sense you formulate a question for the database in the client, you submit it to the database, and you get back an answer. Of course those server processes can run on dedicated and powerful machines, or directly in the cloud, where they are used to work with arbitrarily huge amounts of data. 

We will focus on a relatively recent addition to the database family called [`duckdb`](https://duckdb.org). The main benefit (beyond it's excellent performance) is that it is very easy to run a database _without_ the need to set up the client-server structure. We will come to that in due course.

### What is `SQL`

As already mentioned, this is the main database language. Here is an example:

```sql
SELECT carat, cut, clarity, color, price 
  FROM diamonds 
  WHERE price > 15000
```

This reads: 

> From the table called `diamonds` take the columns `carat, cut, clarity, color, price`, but keep only rows where the column `price` is larger than the value `15000`

Not too bad, right?

Let's get this example running now in our computers. We need to create a `duckdb` database first, read some data into it, and the confront it with that `SQL` query. For simplicity we will do this only in `R` to get us started, then we will move on to a more elaborate example and big data, which will use both `R`, `python` and `julia`.

## Simple Intro Example with `R`

```{r setup}
library(DBI)
library(duckdb)
library(tibble)
con = DBI::dbConnect(duckdb::duckdb(), shutdown = TRUE)  # will get erased after shutdown
# con = DBI::dbConnect(duckdb::duckdb(), dbdir = "db.duck") # would save on disk
```

The `con` object is our live connection to the database. Think of it like a telephone receiver - you can talk into it, and it will respond results back to you. As such, we have to refer to `con` in everything we want to do with the database.

Let us first read some data from the `ggplot2` package, and add it to our database. Remember, a database is a collection of tables, and now we are adding two of those:

```{r add_data}
dbWriteTable(con, "mpg", ggplot2::mpg)
dbWriteTable(con, "diamonds", ggplot2::diamonds)
```

Ok, let's see what tables we have in the database now:

```{r listdb}
dbListTables(con)
```

Great. Now how do get data _out_ of it? Here goes the `mpg` dataset:

```{r getout}
con |>
	dbReadTable("mpg") |>
    as_tibble()
```

Back to our SQL query from above. We can submit it to the database listening at `con`:

```{r query1}
query1 = "SELECT carat, cut, clarity, color, price 
  FROM diamonds 
  WHERE price > 15000"

head(dbGetQuery(con, query1))
```

or of course as before with a pipeline such as

```{r query2}
con |>
	dbGetQuery(query1) |>
	as_tibble()
```


That's it for starters. There is now a whole world out there to learn in terms of SQL queries, database joins, pivots (i.e. reshape operations) etc. I leave you in the hands of Hadley Wickham and his [dedicated chapter](https://r4ds.hadley.nz/databases) on this topic if you want to dive deeper. It will be almost unavoidable to learn some simple SQL, but hopefully it won't be so bad. You will see in the associated pages of this tutorial that there are some very nice translation tools for `R`. Grant McDermott's [intro to databases](https://raw.githack.com/uo-ec607/lectures/master/16-databases/16-databases.html#Databases_101)
is another valuable resource here.

```{r reticulate_config}
#| cache: false
# setup python from R with reticulate package
library(reticulate)
Sys.setenv(RETICULATE_PYTHON=here::here(".venv/bin/python"))

```

## Load libraries

::: {.panel-tabset group="language"}

### R

```{r}
#| cache: false

library(duckdb)  # loaded already 
```

### Python

```{python}
#| cache: false

import duckdb
import time # just for timing some queries
```

:::

## Create a database connection

The first thing we need to do is instantiate a connection with an _in-memory_
database.^[Aside: The `shutdown = TRUE` argument is a convenience feature that
ensures our connection is automatically terminated when our R session ends
(i.e., even if we forget to do it manually.) I'm not aware of a similar
convenience argument for Python; please let me know if I am missing something.]

::: {.panel-tabset group="language"}

### R

```{r db_con}
#| cache: false
 
con = dbConnect(duckdb(), shutdown = TRUE)
```

### Python 

```{python db_con_py}
#| cache: false

con = duckdb.connect(database = ':memory:', read_only = False)
```

:::

#### Digression: In-memory versus on-disk

The fact that our connection lives "in memory" is a killer feature of DuckDB (one
that it inherits from SQLite). We don't need to connect to some complicated,
existing database infrastructure to harness all of DuckDB's power. Instead we can just spin up
an ephemeral database that interacts directly with our R (or Python, or Julia,
etc.) client.

However, it's worth noting that you _can_ create a persistent,
disk-backed database simply by providing a database file path argument as part
of your connection, e.g.

::: {.panel-tabset group="language"}

### R

```r
## Uncomment and run the next line if you'd like to create a persistent,
## disk-backed database instead.

# con = dbConnect(duckdb(), dbdir = "nyc.duck")
```

### Python 

```python
## Uncomment and run the next line if you'd like to create a persistent,
## disk-backed database instead.

# con = duckdb.connect(database = 'nyc.duck', read_only = False)
```

:::


(Note that the `".duck"` file extension above is optional. You could also use `".db"`, `".dbb"`, or whatever you want really.)

:::{.callout-important}
## Bigger than RAM data?
One really important benefit of creating a persistent disk-backed database is
that it enables out-of-core computation for bigger than RAM data. See
[here](https://duckdb.org/2024/03/29/external-aggregation.html) for more details
and performance considerations (which are still great).
:::

## Just How Big is This?

Before we get started with some analysis tasks, here's a simple question to ask our database: how many rows are we dealing with here, actually? 

::: {.panel-tabset group="language"}

### R

```{r count-rows}
nrows = dbGetQuery(
      con,
      "
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        COUNT(passenger_count) as nrows
      "
    )
```


### python

```{python count-rows-py}
# con.
#   query(
# 	'''
#     FROM 'nyc-taxi/**/*.parquet'
#     SELECT
#     COUNT(passenger_count) as nrows
# 	'''
# )
```

:::

Ok, well that is `r nrows$nrows`. 😬


## First example

We'll start with a simple aggregation query to get situated. I'll also use
this example to highlight some general features of DuckDB SQL and
the underlying query engine.

Okay, first query. Let's say we want to know: _What is the average tip per
passenger count?_ A typical SQL job for this question might look as follows:

```sql
SELECT
  passenger_count,
  AVG(tip_amount) AS mean_tip
FROM 'nyc-taxi/**/*.parquet'
GROUP BY passenger_count
ORDER BY passenger_count
```

(Where the last `ORDER BY` statement is optional. Note that ordering (i.e.,
sorting) is a potentially expensive operation but we'll get back to that later.)

This is perfectly valid DuckDB SQL too. However, we can rewrite it with slightly
nicer syntax thanks DuckDB's
"[friendly SQL](https://duckdb.org/docs/guides/sql_features/friendly_sql)". The 
key changes for this simple query are going to be: (1) the `FROM` statement
comes first, and (2) we can use the `GROUP BY ALL` and `ORDER BY ALL` statements to avoid writing out
the "passenger_count" grouping column multiple times.^[I'll admit that the benefits don't
seem so great for this simple example. But trust me: they make a big difference
once you start having lots of grouping columns and complex sub-queries.]

```sql
FROM 'nyc-taxi/**/*.parquet'
SELECT
  passenger_count,
  AVG(tip_amount) AS mean_tip
GROUP BY ALL
ORDER BY ALL
```

:::{.callout-tip}
### DuckDB's "friendly SQL"
One of the under-appreciated (IMHO) features of DuckDB is that it supports many
syntax enhancements over tradional SQL dialects, which they collectively dub
"[friendly SQL](https://duckdb.org/docs/guides/sql_features/friendly_sql)".
Together these syntax enhancements allow you to write much more ergonomic SQL
queries that cut down on duplication and logical inconsistencies.
:::

To run this operation from our R or Python client, simply pass the SQL query as
a string to our connection. Let's use this as a chance to save the result and
time our query too. 

::: {.panel-tabset group="language"}

### R

```{r db_sql_dat1}
tic = Sys.time()
dat1 = dbGetQuery(
  con,
  "
  FROM 'nyc-taxi/**/*.parquet'
  SELECT
    passenger_count,
    AVG(tip_amount) AS mean_tip
  GROUP BY ALL
  ORDER BY ALL
  "
)
toc = Sys.time()

dat1
toc - tic
```

```{r db_sql_dat1_time}
#| include: false
dat1_time = sprintf("%.2f", toc - tic) 
```

### Python

```{python db_sql_dat1_py}
tic = time.time()
dat1 = (
  con.
  query(
    '''
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      passenger_count,
      AVG(tip_amount) AS mean_tip
    GROUP BY ALL
    ORDER BY ALL
    '''
    )
)
toc = time.time()

dat1
# print(f"Time difference of {toc - tic} seconds")
## Timing will be misleading for this rendered Quarto doc, since we're calling
## Python from R (via the reticulate package).
```

Note that we actually get a **polars** DataFrame as a return object. Click the
callout box below to learn more.

::::{.callout-note collapse="true"}
## Result conversion in Python (click to expand)
By default, the `con.query` method that we are using here will return a
**polars** DataFrame object that Python understands "natively" (i.e., has a
print method for and so on). Behind the scenes, this **duckdb** to **polars**
integration relies on the **pyarrow** library being available to our Python
environment, which have already installed for this workshop.

It's also possible return other types of Python objects. For example, you can
use the `.df()` method to coerce to a **pandas** DataFrame instead, among
several other formats like **numpy** arrays. (Details
[here](https://duckdb.org/docs/api/python/overview#result-conversion).) Given
the focus of this workshop, it won't surprise you to hear that I'm going to
stick with the default **polars** conversion.
::::

:::

So that only took `r dat1_time` seconds in this rendered Quarto doc (and will
likely be even faster when you try in an interactive session). To underscore
just how crazy impressive this is, recall that this **includes the time that it
takes to read the data from disk**. I can almost guarantee that the read +
serialization time alone for traditional data wrangling workflows would take
several minutes, and probably crash my laptop RAM. Don't forget that our full
dataset is nearly 200 million rows deep and 30 columns wide.

_Aside: We clearly have a few outlier typos in our dataset. 254 passengers in a
single taxi trip? I don't think so. We'd probably want to filter these out with
a `WHERE` statement if we were doing serious analysis, but I'm just going to
leave them in for this tutorial._


## Aggregation

Let's try out some more aggregation queries. How about a slightly variation on a our first example query, where we (a) add "month" as a second grouping variable, and (b) subset to only the first
three months of the year. 

::: {.panel-tabset group="language"}

### R

```{r db_sql_dat2}
tic = Sys.time()
dat2 = dbGetQuery(
  con,
  "
  FROM 'nyc-taxi/**/*.parquet'
  SELECT
    month,
    passenger_count,
    AVG(tip_amount) AS mean_tip
  WHERE month <= 3
  GROUP BY ALL
  "
    )
toc = Sys.time()

head(dat2)
toc - tic
```

### Python

```{python db_sql_dat_py}
tic = time.time()
dat2 = (
  con.
  query(
    '''
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      month,
      passenger_count,
      AVG(tip_amount) AS mean_tip
    WHERE month <= 3
    GROUP BY ALL
    '''
  )
)
toc = time.time()

dat2
# print(f"Time difference of {toc - tic} seconds")
## Timing will be misleading for this rendered Quarto doc, since we're calling
## Python from R (via the reticulate package).
```

:::

Note that this query completed even faster than the first one, even though we
added another grouping variable. Reason: Subsetting along our Hive-partitioned
parquet dataset allows DuckDB to take shortcuts. We can see this directly by
prepending an `EXPLAIN` statement to our query to reveal the optmized query
plan.

::: {.panel-tabset group="language"}

### R

```{r db_sql_dat2_explain}
dbGetQuery(
  con,
  "
  EXPLAIN
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      month,
      passenger_count,
      AVG(tip_amount) AS mean_tip
    WHERE month <= 3
    GROUP BY ALL
  "
)
```

### Python

```{python db_sql_dat_explain_py}
con.query(
  '''
  EXPLAIN
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      month,
      passenger_count,
      AVG(tip_amount) AS mean_tip
    WHERE month <= 3
    GROUP BY ALL
  '''
)

```

:::

_tl;dr_ DuckDB is able to exploit the month partition of our dataset, so
subsetting means that it can avoid unecessary data ingestion. Similarly, it only
reads in a select group of columns; that's what the "PROJECTION" part of the
plan denotes. If nothing else, the take-home message is that DuckDB only does
what it needs to. Laziness as a virtue!

Here's a final aggregation example, this time including a high-dimensional
grouping column (i.e., "trip_distance") and some additional aggregations.

::: {.panel-tabset group="language"}

### R

```{r db_sql_dat3}
tic = Sys.time()
dat3 = dbGetQuery(
  con,
  "
  FROM 'nyc-taxi/**/*.parquet'
  SELECT
    passenger_count,
    trip_distance,
    AVG(tip_amount) AS mean_tip,
    AVG(fare_amount) AS mean_fare
  GROUP BY ALL
"
)
toc = Sys.time()

nrow(dat3)
head(dat3)
toc - tic
```

### Python

```{python db_sql_dat3_py}
tic = time.time()
dat3 = (
  con.
  query(
    '''
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      passenger_count,
      trip_distance,
      AVG(tip_amount) AS mean_tip,
      AVG(fare_amount) AS mean_fare
    GROUP BY ALL
    '''
    )
)
toc = time.time()

len(dat3)
dat3
# print(f"Time difference of {toc - tic} seconds")
## Timing will be misleading for this rendered Quarto doc, since we're calling
## Python from R (via the reticulate package).
```

:::

## Pivot (reshape)

Let's explore some pivot (reshape) examples, by building off the previous query.

- `UNPIVOT`: wide => long
- `PIVOT`: long => wide

Here I'll use a
[Common Table Expression (CTE)](https://raw.githack.com/uo-ec607/lectures/master/16-databases/16-databases.html#Common_Table_Expressions)
to define a temporary table `tmp_table`, before unpivoting---i.e., reshaping
long---at the
end.

::: {.panel-tabset group="language"}

### R

```{r db_sql_pivot}
dat_long = dbGetQuery(
  con,
  "
  WITH tmp_table AS (
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      passenger_count,
      trip_distance,
      AVG(tip_amount) AS mean_tip,
      AVG(fare_amount) AS mean_fare
    GROUP BY ALL
  )
  UNPIVOT tmp_table
  ON mean_tip, mean_fare
  INTO
    NAME variable
    VALUE amount
  "
)

head(dat_long)
```

### Python

```{python db_sql_pivot_py}
dat_long = (
  con.
  query(
    '''
    WITH tmp_table AS (
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        passenger_count,
        trip_distance,
        AVG(tip_amount) AS mean_tip,
        AVG(fare_amount) AS mean_fare
      GROUP BY ALL
    )
    UNPIVOT tmp_table
    ON mean_tip, mean_fare
    INTO
      NAME variable
      VALUE amount
    '''
  )
)

dat_long
```

:::

Another option would have been to create a new table in memory
and then pivot over that, which segues nicely to...

### Digression: Create new tables

CTEs are a very common, and often efficient, way to implement multi-table
operations in SQL. But, for the record, we can create new tables in DuckDB's
memory cache pretty easily using the `CREATE TABLE` statement.

::: {.panel-tabset group="language"}

### R

Instead of `DBI::dbGetQuery`, we must now use `DBI::dbExecute`.

```{r, db_sql_taxi2}
dbExecute(
    con,
    "
    CREATE TABLE taxi2 AS
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        passenger_count,
        trip_distance,
        AVG(tip_amount) AS mean_tip,
        AVG(fare_amount) AS mean_fare
      GROUP BY ALL
    "
)

dbListTables(con)
```

FWIW, you can always remove a table with `dbRemoveTable()`.

### Python 

Instead of `con.query()`, we must now use `con.execute()`.

```{python, db_sql_taxi2_py}
con.execute(
  '''
  CREATE TABLE taxi2 AS
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      passenger_count,
      trip_distance,
      AVG(tip_amount) AS mean_tip,
      AVG(fare_amount) AS mean_fare
    GROUP BY ALL
  '''
)

# https://stackoverflow.com/q/75727685
con.query(
  '''
  SELECT table_name, estimated_size AS nrows, column_count AS ncols
  FROM duckdb_tables;
  '''
)
```
:::

### Back to reshaping

With our new `taxi2` table in hand, let's redo the previous unpivot query
directly on this new table. This makes the actual (un)pivot statement a bit
clearer... and also separates out the execution time.

::: {.panel-tabset group="language"}

### R

```{r db_sql_pivot2}
dbGetQuery(
  con,
  "
  UNPIVOT taxi2
  ON mean_tip, mean_fare
  INTO
    NAME variable
    VALUE amount
  LIMIT 5
  "
)
```

### Python

```{python db_sql_ldat2_py}
con.query(
  '''
  UNPIVOT taxi2
  ON mean_tip, mean_fare
  INTO
    NAME variable
    VALUE amount
  LIMIT 5
  '''
)
```

:::

(Note how crazy fast pivoting in DuckDB actually is.)

## Joins (merges)

It's a bit hard to demonstrate a join with only a single main table. But here is
a contrived example, where we calculate the mean monthly tips and the mean
monthly fares as separate sub-queries (CTEs), before joining them together by
month.

::: {.panel-tabset group="language"}

### R

```{r db_sql_join}
dbGetQuery(
  con,
  "
  WITH 
    mean_tips AS (
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        month,
        AVG(tip_amount) AS mean_tip
      GROUP BY month
    ),
    mean_fares AS (
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        month,
        AVG(fare_amount) AS mean_fare
      GROUP BY month 
    )
  FROM mean_tips
  LEFT JOIN mean_fares
  USING (month)
  SELECT *
  ORDER BY mean_tips.month
  "
)
```

### Python

```{python db_sql_join_py}
con.query(
  '''
  WITH 
    mean_tips AS (
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        month,
        AVG(tip_amount) AS mean_tip
      GROUP BY month
    ),
    mean_fares AS (
      FROM 'nyc-taxi/**/*.parquet'
      SELECT
        month,
        AVG(fare_amount) AS mean_fare
      GROUP BY month 
    )
  FROM mean_tips
  LEFT JOIN mean_fares
  USING (month)
  SELECT *
  ORDER BY mean_tips.month
  '''
)
```

:::

:::{.callout-tip}
## Challenge

Redo the above join but, rather than using CTEs, use tables that you first
create in DuckDB's memory bank. Again, this will simplify the actual join
operation and also emphasise how crazy fast joins are in DuckDB.
:::

## Windowing

One last example: Binning "trip_distance" into deciles and then calculating the
the mean "tip_amount" within each decile. This is an example of a
[window function](https://duckdb.org/2021/10/13/windowing.html)
and query pattern that I use _all the time_ in my own work. I find it extremely
useful for quickly pulling out descriptive patterns from large datasets, from
which I can then develop a better intuition of my data. In turn, this shapes the
hypotheses and modeling choices that I make in the subsequent analysis stage.

:::{.callout-warning}
## Sorting and sampling
I'm using a 1% random sample of my data here (see the `USING SAMPLE 1%`
statement).  Why? Because calculating deciles requires ranking your data and
this is expensive! To rank data, we first have to sort it (`ORDER`) and this
requires evaluating/comparing every single row in your dataset. In turn, this
means that it's very hard to take shortcuts. (This is one reason why DuckDB's
optimized query plan will always delay sorting until as late as possible; to
only sort on a smaller subset/aggregation of the data if possible.) FWIW,
DuckDB's sorting algorithm is still
[crazy fast](https://duckdb.org/2021/08/27/external-sorting.html).
But for data of this size, and where sorting on the full datset is unavoidable,
I strongly recommend sampling first. Your general insights will almost certainly
remain intact.
:::

::: {.panel-tabset group="language"}

### R

```{r db_sql_bins}
dbGetQuery(
  con,
  "
  WITH trip_deciles AS (
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      tip_amount,
      trip_distance,
      NTILE(10) OVER (ORDER BY trip_distance) AS decile
    USING SAMPLE 1%
  )
  FROM trip_deciles
  SELECT
    decile,
    AVG(trip_distance) AS mean_distance,
    AVG(tip_amount) AS mean_tip
  GROUP BY ALL
  ORDER BY ALL
  "
)
```

### Python

```{python db_sql_bins_py}
con.query(
  '''
  WITH trip_deciles AS (
    FROM 'nyc-taxi/**/*.parquet'
    SELECT
      tip_amount,
      trip_distance,
      NTILE(10) OVER (ORDER BY trip_distance) AS decile
    USING SAMPLE 1%
  )
  FROM trip_deciles
  SELECT
    decile,
    AVG(trip_distance) AS mean_distance,
    AVG(tip_amount) AS mean_tip
  GROUP BY ALL
  ORDER BY ALL
  '''
)
```

:::


## Close connection

::: {.panel-tabset group="language"}

### R

```{r db_con_close}
#| cache: false

dbDisconnect(con)
```

Again, this step isn't strictly necessary since we instantiated our connection
with the `shutdown = TRUE` argument. But it's worth seeing in case you want to
be explicit.

### Python

```{python db_con_close_py}
#| cache: false

con.close()
```

:::
