---
title: Polars from Python and R
subtitle: 'Pro-tip: Just swap `.` (Python) for `$` (R), or vice versa'
execute:
  freeze: auto
  cache: true
---

```{r reticulate_config}
#| cache: false
#| include: false
Sys.setenv(RETICULATE_PYTHON=here::here(".venv/bin/python"))
options(polars.df_knitr_print = "html")
```

## Load libraries

::: {.panel-tabset group="language"}

### Python

```{python pl_libs}
#| cache: false
import polars as pl
import time
import matplotlib
```

### R

```{r pl_libs_r}
#| cache: false
library(polars)
```

:::

## Scan data 

::: {.panel-tabset group="language"}

### Python

```{python pl_nyc}
#| cache: false
nyc = pl.scan_parquet("nyc-taxi/**/*.parquet", hive_partitioning=True)
nyc
```

### R

```{r pl_nyc_r}
#| cache: false
nyc = pl$scan_parquet("nyc-taxi/**/*.parquet", hive_partitioning=TRUE)
nyc
```
:::

## First example

Polars operations are registered as queries until they are collected.

::: {.panel-tabset group="language"}

### Python

```{python pl_q1}
q1 = (
    nyc
    .group_by(["passenger_count"])
    .agg([
            pl.mean("tip_amount")#.alias("mean_tip") ## alias is optional
        ])
    .sort("passenger_count")
)
q1
```

### R

```{r pl_q1_r}
q1 = (
    nyc
    $group_by("passenger_count")
    $agg(
        pl$mean("tip_amount")#$alias("mean_tip") ## alias is optional
    )
    $sort("passenger_count")
)
q1 
```

::::{.callout-note}
## R-polars multiline syntax

Polars-style `x$method1()$method2()...` chaining may seem a little odd to R users, especially for multiline queries. Here I have adopted the same general styling as Python: By enclosing the full query in parentheses `()`, we can start each `$method()` on a new line. If this isn't to your fancy, you could also rewrite these multiline queries as follows:

```r
nyc$group_by(
    "passenger_count"
)$agg(
    pl$mean("tip_amount")
)$sort("passenger_count")
```
::::

:::

(Note: this is the naive query plan, not the optimized query that **polars**
will actually implement for us. We'll come back to this idea shortly.)

Calling `collect()` enforces computation.

::: {.panel-tabset group="language"}

### Python

```{python pl_dat1}
tic = time.time()
dat1 = q1.collect()
toc = time.time()

dat1
# print(f"Time difference of {toc - tic} seconds")
```

### R

```{r pl_dat1_r}
#| dependson: pl_q1_r
tic = Sys.time()
dat1 = q1$collect()
toc = Sys.time()

dat1
toc - tic
```

:::

## Aggregation

Subsetting along partition dimensions allows for even more efficiency gains.

::: {.panel-tabset group="language"}

### Python

```{python pl_q2}
q2 = (
    nyc
    .filter(pl.col("month") <= 3)
    .group_by(["month", "passenger_count"])
    .agg([pl.mean("tip_amount").alias("mean_tip")])
    .sort("passenger_count")
)
```

### R

```{r pl_q2_r}
q2 = (
    nyc
    $filter(pl$col("month") <= 3)
    $group_by("month", "passenger_count")
    $agg(pl$mean("tip_amount")$alias("mean_tip"))
    $sort("passenger_count")
) 
```

:::

Let's take a look at the optimized query that Polars will implement for us.

::: {.panel-tabset group="language"}

### Python

```{python pl_q2_show}
# q2             # naive
q2.show_graph()  # optimized
```

### R 

```{r pl_q2_expain_r}
#| dependson: pl_q2_r
# q2              # naive
cat(q2$explain()) # optimized
```

:::

Now, let's run the query and collect the results.

::: {.panel-tabset group="language"}

### Python

```{python pl_dat2}
tic = time.time()
dat2 = q2.collect()
toc = time.time()

dat2
# print(f"Time difference of {toc - tic} seconds")
```

### R

```{r pl_dat2_r}
#| dependson: pl_q2_r
tic = Sys.time()
dat2 = q2$collect()
toc = Sys.time()

dat2
toc - tic
```

:::

High-dimensional grouping example.
This query provides an example where **polars** is noticeably slower than
DuckDB.

::: {.panel-tabset group="language"}

### Python

```{python pl_dat3}
q3 = (
    nyc
    .group_by(["passenger_count", "trip_distance"])
    .agg([
        pl.mean("tip_amount").alias("mean_tip"),
        pl.mean("fare_amount").alias("mean_fare"),
        ])
    .sort(["passenger_count", "trip_distance"])
)

tic = time.time()
dat3 = q3.collect()
toc = time.time()

dat3
# print(f"Time difference of {toc - tic} seconds")
```

### R

```{r pl_dat3_r}
q3 = (
    nyc
    $group_by("passenger_count", "trip_distance")
    $agg(
        pl$mean("tip_amount")$alias("mean_tip"),
        pl$mean("fare_amount")$alias("mean_fare")
        )
    $sort("passenger_count", "trip_distance")
)

tic = Sys.time()
dat3 = q3$collect()
toc = Sys.time()
 
dat3
toc - tic
```

:::

As an aside, if we didn't care about column aliases (or sorting), then the previous query could be shortened to:

::: {.panel-tabset group="language"}

### Python

```python
(
    nyc
    .group_by(["passenger_count", "trip_distance"])
    .agg(pl.col(["tip_amount", "fare_amount"]).mean())
    .collect()
)
```

### R

```r
(
    nyc
    $group_by("passenger_count", "trip_distance")
    $agg(pl$col("tip_amount", "fare_amount")$mean())
    $collect()
)
```

:::

## Pivot (reshape)

In **polars**, we have two distinct reshape methods:

- `pivot`: => long to wide
- `unpivot`: => wide to long 

Here we'll _unpivot_ to go from wide to long and use the eager execution engine
(i.e., on the `dat3` DataFrame object that we've already computed) for
expediency.

::: {.panel-tabset group="language"}

### Python

```{python pl_unpivot}
#| dependson: pl_dat3
dat3.unpivot(index = ["passenger_count", "trip_distance"])
```

### R

```{r pl_unpivot_r}
#| dependson: pl_dat3_r
dat3$unpivot(index = c("passenger_count", "trip_distance"))
```

:::

## Joins (merges)

::: {.panel-tabset group="language"}

### Python

```{python pl_join0}
mean_tips  = nyc.group_by("month").agg(pl.col("tip_amount").mean())
mean_fares = nyc.group_by("month").agg(pl.col("fare_amount").mean())
```

```{python pl_join1}
(
    mean_tips
    .join(
        mean_fares,
        on = "month",
        how = "left" # default is inner join
    )
    .collect()
)
```

### R

```{r pl_join0_r}
mean_tips  = nyc$group_by("month")$agg(pl$col("tip_amount")$mean())
mean_fares = nyc$group_by("month")$agg(pl$col("fare_amount")$mean())
```

```{r pl_join1_r}
(
    mean_tips
    $join(
        mean_fares,
        on = "month",
        how = "left"  # default is inner join
    )
    $collect()
)
```

:::


## Appendix: Alternate interfaces

The native **polars** API is not the only way to interface with the underlying
computation engine. Here are two alternate approaches that you may prefer,
especially if you don't want to learn a new syntax.

### Ibis (Python)

The great advantage of **Ibis** (like **dbplyr**) is that it supports multiple
backends through an identical frontend. So, all of our syntax logic and workflow
from the Ibis+DuckDB section carry over to an equivalent Ibis+Polars workflow
too. All you need to do is change the connection type. For example:

```{python pl_ibis_example}
import ibis
import ibis.selectors as s
from ibis import _

##! This next line is the only thing that's changed !##
con = ibis.polars.connect()

con.register("nyc-taxi/**/*.parquet", "nyc")

nyc = con.table("nyc")

(
  nyc
  .group_by(["passenger_count"])
  .agg(mean_tip = _.tip_amount.mean())
  .to_polars()
)
```

### tidypolars (R)

The R package **tidypolars** ([link](https://www.tidypolars.etiennebacher.com/)) provides the "tidyverse" syntax while using **polars** as backend. The syntax and workflow should thus be immediately familar to R users.

It's important to note that **tidypolars** is _solely_ focused on the
translation work. This means that you still need to load the main **polars**
library alongside it for the actual computation, as well as **dplyr** (and
potentially **tidyr**) for function generics.

```{r pl_tidypolars_example_r}
library(polars) ## Already loaded
library(tidypolars)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)

nyc = scan_parquet_polars("nyc-taxi/**/*.parquet")

nyc |> 
    summarise(mean_tip = mean(tip_amount), .by = passenger_count) |>
    compute()
```

_Aside: Use `collect()` instead of `compute()` at the end if you would prefer to
return a standard R data.frame instead of a Polars DataFrame._

See also **polarssql** ([link](https://rpolars.github.io/r-polarssql/)) if you
would like yet another "tidyverse"-esque alternative that works through
**DBI**/**d(b)plyr**.
